---
title: "MCnebula workflow for LC-MS/MS dataset analysis"
author: "Lichuang Huang; Lu Wang; Qiyuan Shan; Qiang Lv; Keda Lu; Gang Cao"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette descrip a classified visualization method, called MCnebula, for the analysis of untargeted LC-MS/MS datasets.
MCnebula utilizes the state-of-the-art computer prediction technology, SIRIUS workflow (SIRIUS, ZODIAC, CSI:fingerID, CANOPUS), for compound formula prediction, structure retrieve and classification prediction.
MCnebula integrates an abundance-based class selection algorithm into compound annotation.
The benefits of molecular networking, i.e. intuitive visualization and a large amount of integratable information, were incorporated into MCnebula visualization.
With MCnebula, we can switch from untargeted to targeted analysis, focusing precisely on the compound or chemical class of interest to the researcher.

## R and other softs Setup

### R setup

```{r setup}
library(MCnebula)
library(dplyr)
library(ggplot2)
library(ggraph)
library(grid)
```

### Others setup

The prerequisite soft for MCnebula is SIRIUS 4.
Users can download that from <https://bio.informatik.uni-jena.de/software/sirius/>.  

&ensp;&ensp; 
In addition, users are encouraged to perform MZmine 2 (<https://github.com/mzmine/mzmine2>) for LC-MS/MS data processing.

&ensp;&ensp; 
Before mass spectrometry data been processed, the raw data should be converted to .mzML or .mzXML file in most cases.
ProteoWizard (<https://proteowizard.sourceforge.io/>) could be implemented for conversion.

## Data preprocessing

### Raw data processing

For MZmine2 processing, an XML batch file outlined the example parameters for waters Qtof could be find in <https://github.com/Cao-lab-zcmu/research-supplementary>.

### SIRIUS computation workflow

Here we prepared some example files for this vignette to better illustrate MCnebula workflow.

```{r, echo = T}
eg.path <- system.file("extdata", "raw_instance.tar.gz", package = "MCnebula")
tmp <- tempdir()
utils::untar(eg.path, exdir = tmp)
mgf.path <- paste0(tmp, "/", "instance5.mgf")
### show details of .mgf
data.table::fread(mgf.path, header = F, sep = NULL)
```

If SIRIUS soft has been download locally and the Environment Path has been set down (for example, set `PATH=$PATH:/you/your_dir/sirius-gui/bin` in `~/.bashrc`), the following is availlable:

```{r, eval = F}
system(
  paste0(
    "sirius -i ",
    mgf.path,
    " -o test --maxmz 800 formula -c 50 zodiac structure canopus"
    )
)
```

A more simpler approach is to use SIRIUS GUI version.

### MCnebula processing

In deed, the computation of SIRIUS is time-consumed. It may cost several hours even days.
Here, we prepared a fairly small dataset (SIRIUS project space) which has been done from SIRIUS computation.
The following show the detail of this dataset.

```{r, echo = T}
list.files(tmp) %>% 
  .[which(. != "instance5.mgf")]
```

#### Data collating

To begin with, MCnebula should he initialized in SIRIUS project.

```{r, echo = T}
MCnebula::initialize_mcnebula(tmp, rm_mc.set = T)
```

This will set some global var.

```{r, echo = T}
ls(pattern = "^\\.MCn\\.", all.names = T)
```

Users can manually modify the setup.

##### Collate structure

```{r, echo = T}
MCnebula::collate_structure(
  exclude_element = c("Cl", "S", "P"),
  ppm_error = 20
)
```

The results:

```{r, echo = T}
.MCn.formula_set
.MCn.structure_set
```

Next, we collate the metadata of chemical classes hierarchy.

```{r, echo = T}
MCnebula::build_classes_tree_list()
```

The frame of returned project, e.g.:

```{r, echo = T}
.MCn.class_tree_list[[4]]
```

##### Collate PPCP data.

This function retrieve all posterior probability of classification prediction (PPCP) of compounds and gathered them as `.MCn.ppcp_dataset`.
Furthermore, in default, this function also summarise `.MCn.nebula_class` and `.MCn.nebula_index` upon the data of `.MCn.ppcp_dataset`.
If done, `.MCn.nebula_index` would be returned.

```{r, echo = T, eval = T}
MCnebula::collate_ppcp(
  ## due to the size of example dataset, we set much smaller herein
  ## min_possess, 20 or more may be better
  min_possess = 1,
  ## 0.1 may better
  max_possess_pct = 0.9
)
```

In addition, `collate_ppcp` would set following global var:

```{r, echo = T}
head(.MCn.ppcp_dataset, n = 1)
head(.MCn.nebula_class, n = 1) %>% 
  lapply(dplyr::as_tibble)
```

&ensp;&ensp; 
Indeed, the above prepared instance data is too small to show the detail of the following function of MCnebula.
Hence, we used another projects which has been collated done via above functions.

```{r, echo = T}
### the following project were auto-loaded while use 'library(MCnebula)'
inst_formula_set
inst_structure_set
head(inst_ppcp_dataset, n = 1)
```

We set these data as global var, so as the MCnebula function could recognized them without additional parameters.

```{r, echo = T}
.MCn.formula_set <- inst_formula_set
.MCn.structure_set <- inst_structure_set
.MCn.ppcp_dataset <- inst_ppcp_dataset
```

Re-execute the command:

```{r, echo = T, eval = T}
MCnebula::collate_ppcp(
  min_possess = 20,
  max_possess_pct = 0.1
)
```

#### Generate network graph

##### Generate parent-nebula graph

Of note, this function will conduct spectral similarity (fragmentation spectra) computation and reformated these similarity (cosine) as edges file.
This is performed via `MSnbase::compareSpectra`.
It usually cost some time to get results.
Here, we used a randomly formed edges file to avoid it.

```{r, echo = T}
edges <- .MCn.formula_set$.id %>% 
  combn(2) %>% 
  t()
set.seed(100)
### randomly get combination.
edges <- edges[sample(1:nrow(edges), 1000), ] %>% 
  data.frame() %>% 
  dplyr::as_tibble() %>%
  dplyr::rename(
    .id_1 = 1,
    .id_2 = 2
  ) %>% 
  dplyr::mutate(
    dotproduct = rnorm(1000, mean = 0.5, sd = 0.1), 
    mass_diff = rnorm(1000, mean = 200, sd = 50),
  )
tmp.edges <- tempfile()
### write done
MCnebula::write_tsv(edges, filename = tmp.edges)
### show details
edges
```

Subsequently, we use this edges file to generate parent-nebula graph.

```{r, echo = T}
MCnebula::generate_parent_nebula(
  rm_parent_isolate_nodes = T,
  ## specify the edges file
  edges_file = tmp.edges
)
```

This function will generate nodes data (chemical formula and strucutre annotation) and edges data (if the edges file were not specified but real-time computed).
For most, parent-nebula graph was generated.

```{r, echo = T}
.MCn.parent_nodes
.MCn.parent_graph
```

##### Generate child-nebulae graph

This function generate multiple network graph upon parent-nebula (nodes and edges) according to compound classification.

```{r, echo = T}
MCnebula::generate_child_nebulae(
  max_edges = 5,
  ## this will write a output of .graphml, which support by Cytoscape.
  output_format = "graphml"
)
```

These graph were stored in list.

```
head(.MCn.child_graph_list, n = 2)
```

#### Visualization of chemical-nebula

##### Visualization of parent-nebula

All visualization in MCnebula package are output with .svg image.

```{r, eval = F}
MCnebula::visualize_parent_nebula(
  layout = "kk",
  ## map nodes color with superclass
  nodes_color = c("hierarchy" = 3),
  width = 25,
  height = 20
)
```

`rsvg::rsvg_png` could be applied to convert .svg to other png file.

```{r, eval = F}
from.svg <- paste0(
  .MCn.output, "/", .MCn.results,
  "/parent_nebula/parent_nebula.svg"
)
to.png <- paste0(
  .MCn.output, "/", .MCn.results,
  "/parent_nebula/parent_nebula.png"
)
rsvg::rsvg_png(from.svg, to.png)
```

##### Visualization of child-nebula

This function draw child-nebula in grid panel.
The overview of child-nebula shows the abundant classes of this LC-MS/MS dataset.

```{r, eval = F}
MCnebula::visualize_child_nebulae()
```

## Information

All output files:

```{r, eval = T}
list.files(paste0(.MCn.output, "/", .MCn.results), recursive = T)
```

Other information:

```{r, echo = T}
sessionInfo()
```
